Topic modeling on EU and the USA research projects

### 2017-06-26 thesis_report/draft_thesis.tex
add comments

@@

Nowadays, there are many funding opportunities for researchers. For instance,
European Union countries are encouraged to invest 3\% of their GDP in research
and development by 2020. EU's latest research programme, Horizon 2020, allows
researchers to identify new opportunities and directions in almost any field of
research. The programme's goals are to strengthen the EU's position in science,
support industrial innovation and address social concerns, such as climate
change, green transportation and energy.

Another funding agency that actively promotes the progress of science is the
National Science Foundation in the United States of America. NSF supports
research in all the non medical fields of science \& engineering and aims to
advance the national health and prosperity. It finds approximately 24\% of the
federally supported research conducted by the USA's colleges and universities.

With our work, we aim to investigate the evolution of scientific research in the
EU and the US. In other words, we attempt to identify which subjects are more
likely to get a funding, based on the previous focus the two regions in research
fields. In order to answer these questions, we tried to gain some insights from
the projects that got funded the last two decades under the EU Framework
Programmes and the NSF in USA.

When researchers apply for a funding, one of the first steps they do is to
identify an interesting subject of the many possible topics of scientific
investigation. The topic(s) of a project is always described in the project's
abstract/description. Obtaining these abstracts, one can possibly investigate
the topics of the projects and find the subjects that are being addressed and
gain some information regarding the popularity of the topics among others.

In this work, we discover topics and their evolution from 1994 to 2017, using
the projects' abstracts that got funded this period. Using a topic model, we
extract a set of topics and investigate the changes in the content per
year/framework programme. These topics are used to gain insight into some of the
research that occurs the last two decades. Our goal is to share these insights
with researchers and encourage them to develop better strategies to establish
themselves as more effective researchers globally, by choosing a subject that is
more likely to get a funding.

In section 2, we briefly talk about the papers that mostly influenced our study.
Section 3 introduces some basic concepts in topic modeling and we describe the
model that we used for topic extraction, Latent Dirichlet Allocation, which is a
statistical generative model for documents. In section 4 we provide information
about the dataset structure and the methodology that we followed in our
analysis. In section 5 we interpret the results of our analysis, discuss the
evolution of the topics over time and the differences between the two regions'
topics based on their similarity. Finally, section 6 summarizes our work and
section 7 suggests some future work.

\section{Related work}

This chapter surveys previous work in topic modeling on scientific data. Topic
modeling algorithms provides a suite of algorithms to discover hidden thematic
structure in large collections of texts and can be adapted to many kinds of
data. These algorithms can be used to summarize, visualize and explore a
collection of documents. The papers listed below are applying a Latent Dirichlet
Allocation model in an interesting approach on their corpus.

D.Blei's research focuses on probabilistic topic models, Bayesian methods and
approximate posterior inference. In his work "Introduction to Probabilistic
Topic Models",\cite{Blei11introductionto} he describes the basic ideas behind
LDA and applies the model to articles from the journal "Science". The model
assumes that the topics are generated first, before the documents and that the
documents exhibit multiple topics, but each one of them exhibit those topics in
different proportion. In our implementation we also use the LDA model and we
follow a similar approach, since we don't use any previous information about the
subjects that run through our dataset and the documents are not labeled with
keywords or topics. The topic distributions arise by computing the hidden
structure that resembles the thematic structure of the collection.

@@

In this chapter we will present some basic theoretic background about topic
modeling. To deal with extensive text body, machine learning researchers have
developed topic modeling: a suite of methods that analyze the words of the
original text to discover the themes that run through the data and detect
hidden semantic structures by analyzing a corpus. By marking up the documents
using these topics and by using the resulting structure, the model can be used
for organization, finding similar documents, or any other similar tasks. A
topic is here defined as a probability distribution of words, meaning that
certain words are more likely within a certain topic. Since the topics emerge
from the analysis of the original texts, topic modeling algorithms do not
require any prior annotations or labeling of the documents.\cite{Blei11introductionto}

@@

Some further common concepts and terms in topic modeling: 
\begin{itemize}
\item[] A \textit{word} is defined as an item from a vocabulary indexed from 
${1, ..., V}$, where V is the size of the vocabulary. All words are 
represented as unit-basis vectors with one component equal to one and 
the rest equal to zero.
\item[] A \textit{document} is a collection of words denoted by $w = {w_1, ..., w_N}$, 
where $w_n$ is the $n$th word and $N$ is the total number of words in the 
collection.
\item[] A \textit{corpus} is a collection of documents in a dataset. It is denoted 
$D = {w_1,..., w_M}$, where $w_m$ is the $m$th document in the corpus and 
$M$ is the total number of documents.
\item[] \textit{Latent variables} are variables that may not be directly observed, 
unlike observable variables. Latent variables can instead be inferred 
from other observable variables.
\item[] \textit{Polysemy} is the capacity for a word to have multiple related meanings. 
An example of this is the word “plant” which can mean a living organism of 
the kind exemplified by trees, herbs etc., or a place where an industrial or
manufacturing process takes place.
\item[] \textit{Synonymy} is the capacity for several words to have similar meanings such as
the words “buy” and “purchase”.
\end{itemize}

@@

When introducing the Latent Dirichlet Allocation in section 2.2.3, we 
will be using the terms prior and posterior probabilities and 
likelihood function:

\begin{itemize}
\item The prior probability distribution of and uncertain quantity is the 
probability distribution that express the value of this quantity before 
some evidence is taken into account.
\item The posterior probability is the conditional probability after some 
evidence is taken into account.
\item The likelihood function of a set of parameters given some outcome 
is equal to the probability of observing said outcome given those 
parameter values.
\end{itemize}

@@

In the context of probabilities, the Jensen Inequality is stated as follows: 
If X is a random variable and f a convex function then:

@@

And, if f is a concave function the inequality turns:

@@

\begin{equation}
D_{KL}(P \mid \mid Q) = \int_{-\infty}^{+\infty} p(x) log\frac{p(x)}{q(x)}dx
\end{equation}

@@

\begin{equation}
D_{KL}(P \mid \mid Q) = \mathbb{E}_{p}[log p(x)] - \mathbb{E}_{p}[log q(x)]
\end{equation}

@@

Once we have a representation for individual documents, the natural next step
is to select a model for a set of documents. Given a training set of
documents, we will choose values for the parameters of a probabilistic model
that make the training documents have high probability. Then, given a test
document, we can evaluate its probability according to the model. The higher
this probability is, the more similar the test document is to the training set.

The probability distribution that we use is the multinomial. In probability 
theory, the multinomial distribution is a generalization of the binomial 
distribution. For example, it models the probability of counts for rolling 
a k-sided dice n times. For n independent trials each of which leads to a 
success for exactly one of k categories, with each category having a given 
fixed success probability, the multinomial distribution gives the 
probability of any particular combination of numbers of successes for the 
various categories\footnote{\url{https://en.wikipedia.org/wiki/Multinomial_distribution}}.

@@

The Dirichlet distribution is a prior for the multinomial distribution. This
means that if the prior distribution of the multinomial parameters is Dirichlet
then the posterior distribution is also a Dirichlet distribution (with
parameters different from those of the prior).

The Dirichlet process is a way to model randomness of a probability mass
function (PMF) with unlimited options (e.g. an unlimited amount of dice in a
bag). The process is similar Polya's Urn\footnote{\url{http://www.statisticshowto.com/dirichlet-distribution/}}, only instead of having a set number
of ball colors you have an unlimited amount:

@@

First, we'll introduce some basic models for topic modeling. The Latent Semantic
Indexing, or LSI, was presented by Deerwester et al. in 1990. The model manages
to deal with the problem that multiple terms can refer to the same meaning,
i.e. synonymy. However, it is not as successful regarding polysemy. The reason
is that every term is represented as just one point in the so-called latent
semantic space. Furthermore, a word that can mean two or more different things
is represented as a weighted average of the different meanings.

After LSI was introduced, Hofmann presented the Probabilistic Latent Semantic
Indexing, or PLSI model. PLSI is a topic model where each word in a
document is generated from a single topic which results in that each document
in a corpus can be represented with a topic distribution. Later, in 2003 Blei et
al. presented the Latent Dirichlet Allocation, or LDA. As opposed to PLSI, LDA
is a statistically generative model for documents where each word in a document
can be generated by all topics.

@@

Latent Semantic Indexing was one of the earliest methods for finding
relationships between documents and the words that occur in them.\cite{Deerwester90indexingby} In LSI a
term-document matrix is created by analyzing the corpus, where the rows
correspond to words and columns to documents. Each element in this sparse
matrix describes the number of times a word occurs in a document but this term
count can also be weighted with for instance TF-IDF. Letting each word
represent a dimension in a very high dimensional space, a document can be seen
as a vector with components corresponding to its weighted term counts.
\cite{Salton:1988:TAA:54259.54260}

@@

This is similar to PLSI, except
that in LDA the topic distribution is assumed to have a Dirichlet prior. 
In practice, this results in more reasonable mixtures of topics in a
document. The reduced document description result in PLSI is a vector where
each element describes the mixing proportion for a topic. A limitation with
PLSI is that there is no generative model for these proportions, making it
difficult to handle unseen documents. The Latent Dirichlet Allocation model
tries to solve this limitation by setting a Dirichlet prior on the topic
distribution.\cite{blei2003latent}

@@

The basic idea in LDA is that we define a topic to be a Dirichlet distribution
over a fixed vocabulary. Technically, the model assumes that the topics are
generated first, before the documents.\cite{Blei11introductionto} Now for 
each document in the collection, we generate the words in a two-stage process:

@@

This statistical model reflects the intuition that documents exhibit multiple
topics. Each document exhibits the topics with different proportion (step 1);
each word in each document is drawn from one of the topics (step 2b), where the
selected topic is chosen from the per-document distribution over topics (step
2a). All the documents in the corpus share the same set of topics, but each
document exhibits those topics with different proportion - that's the
distinguishing characteristic of LDA. A more detailed explanation of the model
follows in section 2.2.3.1.

@@

LDA is a probabilistic model of text documents, which assumes a collection of 
K topics. Each topic defines a multinomial distribution over the vocabulary 
and is assumed to have been drawn from a Dirichlet 
$\beta_k \sim Dirichlet(\eta)$.

In LDA, each document is represented by a mixture of the topics, with weight
$\theta_w^2$ for topic $z$ in document $w$. These weight have a Dirichlet prior,
parametrized by a hyper-parameter $\alpha$. Each topic is a probability
distribution over the vocabulary, with probability $\beta^w_z$ for word $w$ in
topic $z$. We define the following variables and notation:
\begin{itemize}
  \item[] $k$ is the number of topics.
  \item[] $V$ is the number of unique words in the vocabulary.
  \item[] $\theta$  is the topic distribution (of length $k$) for a document,
   drawn from a uniform Dirichlet distribution with parameter $\alpha$.
  \item[] $z_{n}$ is a topic assignment for $word w_{n}$, sampled from 
  $p(z_{n} = i|\theta) = \theta_{i}$.
  \item[] \textbf{w}$ = (w_{1}, ... , w_{N})$ is a document with $N$ words.
  \item[] $w_{n}^{i} = 1$ means that the word $w_{n}$ is the $i$-th word 
  of the vocabulary.
  \item[] $\beta$  is a $k \times V$ matrix, where each row $\beta_{i}$ 
  is the multinomial distribution for the $i$-th topic. That is, 
  $\beta_{ij} = p(w^{j} = 1 | z_{j} = i)$.
\end{itemize}

@@

This
posterior can't be computed directly; it is usually approximated using Markov
Chain Monte Carlo (MCMC) or variational inference. Both methods are effective,
but also present significant computational challenges in large data sets
~\cite{onlineLDAvb}. In our implementation, we're using the online variational
LDA model, which is described in section 2.2.3.4.

The key difference in LDA compared to PLSI is that the process outlined above
actually is generative for documents. This means that, given a new document,
one can use the parameters learned previously to estimate the topic
distribution $\theta$ for this new document.

A limitation with the standard LDA model is the fix vocabulary. This means that
new words cannot enter the model and that you have to know a priori what words
will model the corpus effectively. A setting where this could be problematic is
when analyzing data where new words in the form of slang or names emerge over 
time. Another limitation is the fact that one has to state
the number of topics the model should use. This is not a big issue, but could
lead to some of the resulting topics to be insignificant.
\begin{center}
\begin{figure}[h]
\centering
\includegraphics[width=0.6\textwidth]{figs/LDA_standard_model.png}
\caption{Plate notation of the standard LDA model}
\end{figure}
\end{center}

@@

In LDA, our ultimate goal is to find the posterior distribution of latent topic 
variables after observing training data. However, computing this distribution 
is intractable, so we're forced to approximate. One approximation approach is to 
use an optimization method called Variational Bayes. In the original paper it's 
being used a variational Bayes approximation of the posterior distribution;
\cite{blei2003latent} alternative inference techniques use Gibbs sampling and
expectation propagation. The goal of variational inference is to approximate an
intractable probability distribution $p$, with a tractable one $q$, in a way
that makes them as close as possible given a similarity measure. In this thesis,
variational inference is used to approximate the posterior distribution of
latent variables. A new distribution over the hidden variables $q(z, \beta)$
(called the variational distribution) is  introduced. This new distribution has
properties such that it can be efficiently computed.\cite{Fox2011ATO}

@@

The specific updates are:
\begin{equation}
\phi_{d,word_{i},k} \propto e^{E_{q}(log \theta_{d,k})+E_{q}(log\beta_{k,word_{i}})}
\end{equation}
\begin{equation}
\gamma_{d,k} = \alpha + \sum_{word_{i}}\phi_{d,word_{i},k}n_{d,word_{i}}
\end{equation}
\begin{equation}
\lambda_{k,word_{i}}=\eta +\sum_{d}n_{d,word_{i}}\phi_{d,word_{i},k}
\end{equation}
Where $n_{d,word_{i}}$ is the number of occurrences of $word_{i}$ in document $d$.
The algorithm above has constant memory requirements and empirically
converges faster than collapsed Gibbs sampling. However, it still requires a
full pass through the entire corpus in each iteration. It can be slow to apply
to very large datasets.

@@

The online E-Step only uses the current chunk; instead of holding all the 
documents we now only have to hold 1000 in memory. The E-Step finds locally 
optimal values for $\phi$ and $\gamma$, while in the M-Step, we first compute 
$\lambda^{*}$, which is the value of $\lambda$ if we imagined that the entire 
dataset is made up of $\frac{numDocs}{chunkSize}$ copies of the current chunk. 
Then $\lambda$ is updated using a weighted sum of $\lambda^{*}$ and $\lambda$:

@@

In this section, we explain the procedure that we followed in n order to apply
the LDA model in our dataset. The dataset consists of two datasets: the one that
contains all the information about EU projects founded for research and
innovation under the FP4-H2020 and the dataset from USA about the
awards/projects that got approved from 1994-2017. We consider the projects'
description/abstract as the documents in LDA.

@@

The data that are provided by CORDIS (via the EU Open Data Portal) are open,
free and can be used in research, applications, commercial or non-commercial
purposes. The dataset that we chose for the following analysis contains
information about all the projects funded by the European Union from 1994 to
2017. These projects got approved under the framework programme (FP4-H2020) for
research and technological development. For each project it is provided
information, such as: reference, acronym, dates, funding, programs,
participant countries, subjects and objectives.

In the provided dataset, the information is splitted into CSV files, based on
the framework program that each project belongs. The frameworks from 1994 to
2020 are listed below:

@@

To obtain the collection of the documents, we consider as 'documents' in LDA the
project objectives, which is an accurate description for the project. We also
added the 'title' value of each project to the objectives-documents, to increase
the probability of finding accurate topics. For the majority of the projects the
'start date' and 'end date' value was null, so we kept the splitting of the
dataset per framework, in order to keep the correlation to the time. We followed
the same format in the dataset of USA awards.

@@

As stopwords, we used the list that is provided by the Gensim Python
library\footnote{http://radimrehurek.com/gensim/} and we added following very
common words in the existing dataset: 'data', 'based', 'new', 'project',
'university', 'student', 'students', 'research', 'study', 'program',
'development', 'study', 'studies', 'provide', 'use'. We also removed the rare
words that appear less than 5 times in the collection. Finally, we obtained a
smaller corpus per framework (see the table above)

In this section, we explain the procedure that we followed in n order to apply
the LDA model in our dataset. The dataset consists of two datasets: the one that
contains all the information about EU projects founded for research and
innovation under the FP4-H2020 and the dataset from USA about the
awards/projects that got approved from 1994-2017. We consider the projects'
description/abstract as the documents in LDA.

@@

The National Science Foundation (NSF) is a United States government agency that
supports fundamental research and education in all the non-medical fields of
science and engineering. Its medical counterpart is the National Institutes of
Health. The NSF funds approximately 24\% of all federally supported basic
research conducted by the United States' colleges and universities. In some
fields, such as mathematics, computer science, economics, and the social
sciences, the NSF is the major source of federal backing.

>>>>

\usepackage[textsize=tiny, textwidth=30mm]{todonotes}

\usepackage{geometry}
 \geometry{
 a4paper,
 right=40mm,
 top=20mm,
 }

@@

Nowadays, there are many \todo{ That sounds like a joke. Research
  funding is extremely competitive. } funding opportunities for
researchers. For instance, European Union countries are encouraged to
invest 3\% of their GDP in research and development by 2020. EU's
latest research programme, Horizon 2020, allows researchers to
identify new opportunities and directions in almost any field of
research. The programme's goals are to strengthen the EU's position in
science, support industrial innovation and address social concerns,
such as climate change, green transportation and energy.

Another funding agency that actively promotes the progress of science
is the National Science Foundation in the United States of America.
NSF supports research in all the non medical fields of science \&
engineering and aims to advance the national health and prosperity. It
finds \todo{funds} approximately 24\% \todo{Where did you find this
  number? Add reference.} of the federally supported research
conducted by the USA's colleges and universities.

With our work, we aim to investigate the evolution of scientific
research in the EU and the US. In other words, we attempt to identify
which subjects are more likely to get a funding, based on the previous
focus the two regions in research fields \todo{Does not read right}.
In order to answer these questions, we tried to gain some insights
from the projects that got funded the last two decades under the EU
Framework Programmes and the NSF in USA.

When researchers apply for funding, one of the first steps they do is
to identify an interesting subject of the many possible topics of
scientific investigation. The topic(s) of a project is always
described in the project's abstract/description. Obtaining
\todo{Using} these abstracts, one can possibly investigate the topics
of the projects and find the subjects that are being addressed and
gain some information regarding the popularity of the topics among
others.

In this work, we discover topics and their evolution from 1994 to
2017, using the projects' abstracts that got funded \todo{in} this
period. Using a topic model, we extract a set of topics and
investigate the changes in the content per year/framework programme.
These topics are used to gain insight into some of the research that
occurs the last two decades. Our goal is to share these insights with
researchers and encourage them to develop better strategies to
establish themselves as more effective researchers globally, by
choosing a subject that is more likely to get a funding.

In section 2 \todo{Throughout: never use hard-coded references. Always
  use LaTeX cross-references.}, we briefly talk about the papers that
mostly influenced our study. Section 3 introduces some basic concepts
in topic modeling and we describe the model that we used for topic
extraction, Latent Dirichlet Allocation, which is a statistical
generative model for documents. In section 4 we provide information
about the dataset structure and the methodology that we followed in
our analysis. In section 5 we interpret the results of our analysis,
\todo{and we}discuss the evolution of the topics over time and the
differences between the two regions' topics based on their similarity.
Finally, section 6 summarizes our work and section 7 suggests some
future work.

\section{Related work}\todo{Having this section here, before any
  introduction to LDA, makes counter-intuitive reading. Better move it
after.}

This chapter \todo{section} surveys previous work in topic modeling on
scientific data. Topic modeling algorithms provides \todo{provide} a
suite of algorithms to discover hidden thematic structure
\todo{structures} in large collections of texts and can be adapted to
many kinds of data. These algorithms can be used to summarize,
visualize and explore a collection of documents. The papers listed
below are applying a Latent Dirichlet Allocation model in an
interesting approach on their corpus.

D.Blei's \todo{Throughout: separate initials with space from surname}
research focuses on probabilistic topic models, Bayesian methods and
approximate posterior inference. In his work "Introduction to
Probabilistic Topic Models",\cite{Blei11introductionto}
\todo{Throughout: Fix quotation marks; also citations are placed
  before punctuation} he describes the basic ideas behind LDA and
applies the model to articles from the journal "Science"
\todo{\emph{Science}}. The model assumes that the topics are generated
first, before the documents and that the documents exhibit multiple
topics, but each one of them exhibit those topics in different
proportion. In our implementation we also use the LDA model and we
follow a similar approach, since we don't use any previous information
about the subjects that run through our dataset and the documents are
not labeled with keywords or topics. The topic distributions arise by
computing the hidden structure that resembles the thematic structure
of the collection.

@@

In this chapter we will present some basic theoretic background about
topic modeling. To deal with extensive text body \todo{probably
  plural}, machine learning researchers have developed topic modeling:
a suite of methods that analyze the words of the original text to
discover the themes that run through the data and detect hidden
semantic structures by analyzing a corpus. By marking up the documents
using these topics and by using the resulting structure, the model can
be used for organization, finding similar documents, or any other
similar tasks. A topic is here defined as a probability distribution
of words, meaning that certain words are more likely within a certain
topic. Since the topics emerge from the analysis of the original
texts, topic modeling algorithms do not require any prior annotations
or labeling of the documents.\cite{Blei11introductionto}

@@

Some further common concepts and terms in topic modeling: 
\begin{itemize}
\item[] A \textit{word} is defined as an item from a vocabulary
  indexed from ${1, ..., V}$, where V \todo{$V$} is the size of the
  vocabulary. All words are represented as unit-basis vectors with one
  component equal to one and the rest equal to zero.
\item[] A \textit{document} is a collection of words denoted by
  $w = {w_1, ..., w_N}$, where $w_n$ is the $n$th word and $N$ is the
  total number of words in the collection.
\item[] A \textit{corpus} is a collection of documents in a dataset.
  It is denoted $D = {w_1,..., w_M}$, where $w_m$ is the $m$th
  document in the corpus and $M$ is the total number of documents.
\item[] \textit{Latent variables} are variables that may not be directly observed, 
unlike observable variables. Latent variables can instead be inferred 
from other observable variables.
\item[] \textit{Polysemy} is the capacity for a word to have multiple related meanings. 
An example of this is the word “plant” which can mean a living organism of 
the kind exemplified by trees, herbs etc., or a place where an industrial or
manufacturing process takes place.
\item[] \textit{Synonymy} is the capacity for several words to have similar meanings such as
the words “buy” and “purchase”.
\end{itemize}

@@

When introducing the Latent Dirichlet Allocation in section 2.2.3
\todo{Cross reference}
\todo{As you can see this is a disaster: always cross-reference with
  LaTeX}, we will be using the terms prior and posterior probabilities
and likelihood function:

\begin{itemize}
\item The prior probability distribution of and \todo{an} uncertain
  quantity is the probability distribution that express the value of
  this quantity before some evidence is taken into account.
\item The posterior probability is the conditional probability after some 
evidence is taken into account.
\item The likelihood function of a set of parameters given some outcome 
is equal to the probability of observing said outcome given those 
parameter values.
\end{itemize}

@@

In the context of probabilities, the Jensen Inequality is stated as
follows: If X \todo{$X$} is a random variable and f \todo{$f$} a
convex function then \todo{What is $\mathbb{E}$?}:

@@

And, if f \todo{f} is a concave function the inequality turns:

@@

\begin{equation}
D_{KL}(P \mid \mid Q) = \int_{-\infty}^{+\infty} p(x) log\frac{p(x)}{q(x)}dx
\end{equation} \todo{ log in mathematics is $\log$}

@@

\begin{equation}
  D_{KL}(P \mid \mid Q) = \mathbb{E}_{p}[log p(x)] - \mathbb{E}_{p}[log q(x)]
\end{equation} \todo{Smae with log; also, it is a function.}

@@

Once we have a representation for individual documents, the natural
next step is to select a model for a set of documents. Given a
training set of documents, we will choose values for the parameters of
a probabilistic model that make the training documents have high
probability. Then, given a test document, we can evaluate its
probability according to the model. The higher this probability is,
the more similar the test document is to the training set. \todo{Is
  this really what you use the multinomial distribution for? You use
  the multinomial distribution because LDA assumes that this is the
  distribution of topics in documents and words in topics.}

The probability distribution that we use is the multinomial. In
probability theory, the multinomial distribution is a generalization
of the binomial distribution. For example, it models the probability
of counts for rolling a k-sided dice n times. For n independent trials
each of which leads to a success for exactly one of k categories, with
each category having a given fixed success probability, the
multinomial distribution gives the probability of any particular
combination of numbers of successes for the various
categories\footnote{\url{https://en.wikipedia.org/wiki/Multinomial_distribution}}.
\todo{Variables like $k$, $n$, etc. should be in math mode.}

@@

The Dirichlet distribution is a prior for the multinomial
distribution. This means that if the prior distribution of the
multinomial parameters is Dirichlet then the posterior distribution is
also a Dirichlet distribution (with parameters different from those of
the prior). \todo{ If the Dirichlet is the prior for the multinomial
  then the posterior is the multinomial; in fact the Dirichlet is the
  conjugate prior of the multinomial. It is the posterior distribution
  of the \emph{parameters} that is still Dirichlet. }

The Dirichlet process is a way to model randomness of a probability
mass function (PMF) with unlimited options (e.g. an unlimited amount
of dice in a bag). The process is similar Polya's
Urn\footnote{\url{http://www.statisticshowto.com/dirichlet-distribution/}},
only instead of having a set number of ball colors you have an
unlimited amount: \todo{Why do we need to know that? }

@@

First, we'll introduce some basic models for topic modeling. The
Latent Semantic Indexing, or LSI, was presented by Deerwester et al.
in 1990. \todo{Add reference.} The model manages to deal with the
problem that multiple terms can refer to the same meaning, i.e.
synonymy. However, it is not as successful regarding polysemy. The
reason is that every term is represented as just one point in the
so-called latent semantic space. Furthermore, a word that can mean two
or more different things is represented as a weighted average of the
different meanings.

After LSI was introduced, Hofmann presented the Probabilistic Latent
Semantic Indexing, or PLSI model \todo{Add reference}. PLSI is a topic
model where each word in a document is generated from a single topic
which results in that each document in a corpus can be represented
with a topic distribution. Later, in 2003 Blei et al. presented the
Latent Dirichlet Allocation, or LDA \todo{Add reference}. As opposed
to PLSI, LDA is a statistically generative model for documents where
each word in a document can be generated by all topics.

@@

Latent Semantic Indexing was one of the earliest methods for finding
relationships between documents and the words that occur in
them.\cite{Deerwester90indexingby} In LSI a term-document matrix is
created by analyzing the corpus, where the rows correspond to words
and columns to documents. Each element in this sparse matrix describes
the number of times a word occurs in a document but this term count
can also be weighted with for instance TF-IDF \todo{Explain}. Letting
each word represent a dimension in a very high dimensional space, a
document can be seen as a vector with components corresponding to its
weighted term counts. \cite{Salton:1988:TAA:54259.54260}

@@

This is similar to PLSI, except
that in LDA the topic distribution is assumed to have a Dirichlet prior. 
In practice, this results in more reasonable mixtures of topics in a
document. The reduced document description result in PLSI is a vector where
each element describes the mixing proportion for a topic. A limitation with
PLSI is that there is no generative model for these proportions, making it
difficult to handle unseen documents. The Latent Dirichlet Allocation model
tries to solve this limitation by setting a Dirichlet prior on the topic
distribution.\cite{blei2003latent} \todo{In essence you repeat what
  you said in the beginning of the paragraph.}

@@

The basic idea in LDA is that we define a topic to be a Dirichlet
distribution over a fixed vocabulary. \todo{No, a topic is a
  multinomial distribution.} Technically, the model assumes that the
topics are generated first, before the
documents.\cite{Blei11introductionto} Now for each document in the
collection, we generate the words in a two-stage process:

@@

This statistical model reflects the intuition that documents exhibit
multiple topics. Each document exhibits the topics with different
proportion (step 1); each word in each document is drawn from one of
the topics (step 2b), where the selected topic is chosen from the
per-document distribution over topics (step 2a). All the documents in
the corpus share the same set of topics, but each document exhibits
those topics with different proportion - that's the distinguishing
characteristic of LDA. A more detailed explanation of the model
follows in section 2.2.3.1. \todo{Again, fix cross-references}

@@

LDA is a probabilistic model of text documents, which assumes a
collection of K topics. Each topic defines a multinomial distribution
over the vocabulary and is assumed to have been drawn from a Dirichlet
$\beta_k \sim Dirichlet(\eta)$. \todo{What is $beta_k$ and what is
  $\eta$? If you explain them later (which you do) this sentence
  should follow the explanations.}

In LDA, each document is represented by a mixture of the topics, with
weight $\theta_w^2$ \todo{Where does the square come from?} for topic
$z$ in document $w$. These weight \todo{weights} have a Dirichlet
prior, parametrized by a hyper-parameter $\alpha$. Each topic is a
probability distribution over the vocabulary, with probability
$\beta^w_z$ for word $w$ in topic $z$. We define the following
variables and notation:
\begin{itemize}
  \item[] $k$ is the number of topics.
  \item[] $V$ is the number of unique words in the vocabulary.
  \item[] $\theta$  is the topic distribution (of length $k$) for a document,
   drawn from a uniform Dirichlet distribution with parameter $\alpha$.
 \item[] $z_{n}$ is a topic assignment for $word w_{n}$ \todo{word
     $w_n$}, sampled from $p(z_{n} = i|\theta) = \theta_{i}$.
  \item[] \textbf{w}$ = (w_{1}, ... , w_{N})$ is a document with $N$ words.
  \item[] $w_{n}^{i} = 1$ means that the word $w_{n}$ is the $i$-th word 
  of the vocabulary.
  \item[] $\beta$  is a $k \times V$ matrix, where each row $\beta_{i}$ 
  is the multinomial distribution for the $i$-th topic. That is, 
  $\beta_{ij} = p(w^{j} = 1 | z_{j} = i)$.
\end{itemize}

@@

This
posterior can't be computed directly; it is usually approximated using Markov
Chain Monte Carlo (MCMC) or variational inference. Both methods are effective,
but also present significant computational challenges in large data sets
~\cite{onlineLDAvb}. In our implementation, we're using the online variational
LDA model, which is described in section 2.2.3.4. \todo{Cross reference.}

The key difference in LDA compared to PLSI is that the process outlined above
actually is generative for documents. This means that, given a new document,
one can use the parameters learned previously to estimate the topic
distribution $\theta$ for this new document. \todo{You have already
  said that many times.}

A limitation with the standard LDA model is the fix \todo{fixed}
vocabulary. This means that new words cannot enter the model and that
you have to know a priori what words will model the corpus
effectively. A setting where this could be problematic is when
analyzing data where new words in the form of slang or names emerge
over time. Another limitation is the fact that one has to state the
number of topics the model should use. This is not a big issue, but
could lead to some of the resulting topics to be insignificant.
\begin{center}
\begin{figure}[h]
\centering
\includegraphics[width=0.6\textwidth]{figs/LDA_standard_model.png}
\caption{Plate notation of the standard LDA model}
\end{figure}
\end{center}

@@

In LDA, our ultimate goal is to find the posterior distribution of
latent topic variables after observing training data. However,
computing this distribution is intractable, so we're forced to
approximate. One approximation approach is to use an optimization
method called Variational Bayes. In the original paper it's being used
a variational Bayes approximation of the posterior distribution;
\cite{blei2003latent} alternative inference techniques use Gibbs
sampling and expectation propagation \todo{Cite these techniques.
  AFAIK, if you do not use Variational Bayes, you use MCMC with Gibbs
  sampling.}. The goal of variational
inference is to approximate an intractable probability distribution
$p$, with a tractable one $q$, in a way that makes them as close as
possible given a similarity measure. In this thesis, variational
inference is used to approximate the posterior distribution of latent
variables. A new distribution over the hidden variables $q(z, \beta)$
(called the variational distribution) is introduced. This new
distribution has properties such that it can be efficiently
computed.\cite{Fox2011ATO}

@@

update $\lambda$
\end{algorithm}
\todo{Update $\phi$ and $\gamma$ do not show as separate lines.}

The specific updates are:
\begin{equation}
\phi_{d,word_{i},k} \propto e^{E_{q}(log \theta_{d,k})+E_{q}(log\beta_{k,word_{i}})}
\end{equation}
\begin{equation}
\gamma_{d,k} = \alpha + \sum_{word_{i}}\phi_{d,word_{i},k}n_{d,word_{i}}
\end{equation}
\begin{equation}
\lambda_{k,word_{i}}=\eta +\sum_{d}n_{d,word_{i}}\phi_{d,word_{i},k}
\end{equation}
Where $n_{d,word_{i}}$ is the number of occurrences of $word_{i}$ in
document $d$. The algorithm above has constant memory requirements and
empirically converges faster than collapsed Gibbs sampling. However,
it still requires a full pass through the entire corpus in each
iteration. It can be slow to apply to very large datasets.
\todo{Provide citation for the description you have given; I assume
  it's the same blog post as for the following paragraph.}

@@

The online E-Step only uses the current chunk; instead of holding all the 
documents we now only have to hold 1000 in memory. The E-Step finds locally 
optimal values for $\phi$ and $\gamma$, while in the M-Step, we first compute 
$\lambda^{*}$, which is the value of $\lambda$ if we imagined that the entire 
dataset is made up of $\frac{numDocs}{chunkSize}$ copies of the current chunk. 
Then $\lambda$ is updated using a weighted sum of $\lambda^{*}$ and
$\lambda$: \todo{As the algorithm is thrown on the next page, it looks
  as if something is missing here.}

@@

In this section, we explain the procedure that we followed in order to
apply the LDA model in our dataset. The dataset consists of two
datasets \todo{That reads funny.}: the one that contains all the
information about EU projects founded for research and innovation
under the FP4-H2020 \todo{Throughout: use dashes correctly. There are
  hyphens (-), n-dashes (--) and m-dashes (---).} and the dataset from
USA about the awards/projects that got approved from 1994-2017. We
consider the projects' description/abstract as the documents in LDA.

@@

The data that are provided by CORDIS (via the EU Open Data Portal) are
open, free and can be used in research, applications, commercial or
non-commercial purposes. The dataset that we chose for the following
analysis contains information about all the projects funded by the
European Union from 1994 to 2017. These projects got approved under
the framework programme (FP4-H2020) for research and technological
development. For each project it is provided information \todo{Does
  not read right}, such as: reference, acronym, dates, funding,
programs, participant countries, subjects and objectives.

In the provided dataset, the information is splitted \todo{split} into
CSV files, based on the framework program that each project belongs.
The frameworks from 1994 to 2020 are listed below:

@@

To obtain the collection of the documents, we consider as 'documents'
\todo{In the rest of the document you are using single quotation marks.}
in LDA the project objectives, which is an accurate description for
the project. We also added the 'title' value of each project to the
objectives-documents, to increase the probability of finding accurate
topics. For the majority of the projects the 'start date' and 'end
date' value was null, so we kept the splitting of the dataset per
framework, in order to keep the correlation to the time. We followed
the same format in the dataset of USA awards.

@@

As stopwords, we used the list that is provided by the Gensim Python
library\footnote{http://radimrehurek.com/gensim/} and we added following very
common words in the existing dataset: 'data', 'based', 'new', 'project',
'university', 'student', 'students', 'research', 'study', 'program',
'development', 'study', 'studies', 'provide', 'use'. We also removed the rare
words that appear less than 5 times in the collection. Finally, we obtained a
smaller corpus per framework (see the table above).

In this section, we explain the procedure that we followed in n order
to apply the LDA model in our dataset. The dataset consists of two
datasets: the one that contains all the information about EU projects
founded for research and innovation under the FP4-H2020 and the
dataset from USA about the awards/projects that got approved from
1994-2017. We consider the projects' description/abstract as the
documents in LDA. \todo{Why is this paragraph here?}

@@

The National Science Foundation (NSF) is a United States government
agency that supports fundamental research and education in all the
non-medical fields of science and engineering. Its medical counterpart
is the National Institutes of Health \todo{Does your data include
  material from NIH as well?}. The NSF funds approximately 24\% of all
federally supported basic research conducted by the United States'
colleges and universities. In some fields, such as mathematics,
computer science, economics, and the social sciences, the NSF is the
major source of federal backing.
